---
title: "Assignment 3"
author: "Andrew Wong"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: '3'
  html_document:
    toc: true
    toc_depth: 3
    fig_caption: true
bibliography: "assignment-3.bib"
csl: "biomed-central.csl"
---


# Introduction

## Summary of Dataset and Experimental Design  
## Assignment 1
```{r child="../Assignment-1/assignment-1.Rmd"}
```

## Assignment 2
```{r child="../Assignment-2/assignment-2.Rmd"}
```

## Basic Statistics and Data Source (e.g., GEO ID)

# Non-thresholded Gene Set Enrichment Analysis

## Methods and Gene Sets Used  
## Summary of Enrichment Results  
## Comparison to Thresholded Analysis (Assignment 2)  

# Visualization of Enrichment Results in Cytoscape

## Enrichment Map Creation  
### Node and Edge Counts  
### Thresholds Used  
### Screenshot of Initial Network  

## Network Annotation  
### Parameters Used for Annotation  

## Publication-ready Figure  
### Figure and Legend  

## Theme Network (Collapsed View)  
### Major Themes  
### Fit with the Model  
### Novel Pathways or Themes  

# Interpretation and Detailed View of Results

## Comparison to Original Paper  
## Comparison to Thresholded Results (Assignment 2)  
## Literature Support for Observed Pathways  

# Post-Analysis Exploration

## Option 1: Transcription Factors / microRNAs / Drugs  
### Rationale and Supporting Publications  
### Results of Post-analysis  

## Option 2: Detailed Exploration of a Specific Pathway or Theme  
### Justification for Choice  
### Network or Pathway Diagram  
### Annotation with Fold Change and p-values  

## Option 3: Dark Matter Genes  
### Heatmap: Genes Not Annotated in Returned Pathways  
### Heatmap: Genes Not Annotated in Entire Pathway Set  

# References

